Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-01-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: fleshing out text now we have results

We want to use the Roofline Model to establish the efficiency (or
otherwise) of two shallow-water benchmark codes that we have used in
the GOcean project.

Although we have investigated how the performance of the \psykal
version of nemolite2d compares with that or the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound.

@@

@@

Interpreting the results of the STREAM benchmark in order to get
bandwidth figures for all levels of the cache hierarchy is not
straightforward. We therefore used the STREAM2 benchmark which is
intended for this purpose. In Figure~\ref{FIG_stream2} we plot
the results obtained from the DAXPY component of the STREAM2 benchmark
on the E5-1620 CPU. The plateaux in bandwidth for L2, L3 and main
memory are clear but that for L1 is considerably noisier. The final
values we have used in the roofline plots are given in
Table~\ref{TAB_stream_linpack}.

@@

\begin{table}
\begin{tabular}{|l|c|c|c|c|}
\hline
  \multicolumn{5}{|c|}{E5-1620 v.~2, 3.7GHz} \\
  \hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{28.32} \\
\hline
 &  L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB   & 256 KB   & 10 MB    &  --         \\
\hline
STREAM2 (GB/s)   &  117  &  65.8  & 46.7 &  20.5 \\
  \hline
\multicolumn{5}{|c|}{E5-2697 v.~2, 2.7GHz}  \\
\hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{25.41} \\
\hline
 & L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB  & 256 KB   & 30 MB    &  --         \\
\hline
STREAM2 (GB/s) & 108 & 54.0 & 37.7  &  17.6 \\
\hline
\end{tabular}
\caption{The STREAM and LINPACK benchmark results for a single thread
  running on each of the two CPUs considered here.}
\label{TAB_stream_linpack}
\end{table}

@@

The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
access to Hardware Performance Counters (HWPCs) on Intel and AMD CPUs.
However, there are issues with the accuracy of these counters,
especially for FLOP-related measurements on Ivybridge architectures. A
tool is included in the likwid distribution that tests the accuracy of
HWPC-derived data by running micro-benchmarks for which the FLOP count
and memory traffic are known:
\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
footprint spills out of L1 cache (which all of our nemolite2d
benchmark configurations do), the performance counter results
significantly overestimate the FLOPs performed.

@@

Although the Continuity kernel is very simple, the u-Momentum kernel
is not and in particular, includes a call to the {\tt sin()} function.
In order to estimate the cost of this function in FLOPs the following
code was compiled (with the Intel compiler and -O1) and run with
likwid-perfctr:
\begin{verbatim}
  my_sum = 0.0d0
  do i = 1, 1000
    my_sum = my_sum + sin(real(i))
  end do
  write (*,*) 'My sum = ', my_sum
\end{verbatim}
This measurement revealed that the {\tt sin()} function requires
approximately 13 FLOPs.

A key component of the Roofline model is the Arithmetic Intensity of
the code being executed:
\begin{equation}
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\end{equation}
We calculated this quantity manually by examining the
source code and counting the number of memory references and
arithmetic operations that it contained. In doing this counting we
assume that any references to adjacent array elements are fetched in a
single cache-line and thus only count once.

To illustrate this, we consider the first loop-nest in the
time-stepping section of the Shallow code. This is a doubly-nested
loop over {\tt i} and {\tt j} with {\tt i} innermost). The body of the
loop is:
\begin{verbatim}
CU(I+1,J) = .5d0*(P(I+1,J)+P(I,J))*U(I+1,J)
CV(I,J+1) = .5d0*(P(I,J+1)+P(I,J))*V(I,J+1)
Z(I+1,J+1) =(FSDX*(V(I+1,J+1)-V(I,J+1))- &
             FSDY*(U(I+1,J+1)-U(I+1,J)))/&
            (P(I,J)+P(I+1,J)+            &
             P(I+1,J+1)+P(I,J+1))
H(I,J) = P(I,J)+0.25d0*                   &
         (U(I+1,J)*U(I+1,J)+U(I,J)*U(I,J) & 
         +V(I,J+1)*V(I,J+1)+V(I,J)*V(I,J))
\end{verbatim}
This code writes to four memory locations ({\tt CU}, {\tt CV}, {\tt Z}
and {\tt H}) and reads from ten distinct locations.  However, if we
assume caching of reads from adjacent locations ({\it e.g.} {\tt
  P(i,j)} and {\tt P(i+1,j)}) then we have only six distinct reads.
If we assume that each of these ten memory accesses will touch a
separate cache line then this code fragment will result in the moving
of ten cache lines, each of which is 64 bytes (eight double-precision
words). However, as noted above, this code fragment is the body of a
loop nest. Therefore, upon the next loop trip {\tt i} will have been
incremented by one and the majority of the accessed values will still
be in cache. In fact, a new cache line will only have to be fetched
every seven trips of the (innermost) loop and thus the mean memory
traffic per loop trip is approximately $640/7$ bytes.

We can also count the number of floating-point arithmetic operations
(FLOPs) in this code fragment which gives us 12 multiplications and 12
additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
kernel of nemolite2d then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to
sin(). There are 31 distinct memory references. This gives an $AI$ of
approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we assume
the call to {\tt sin()} costs 13 FLOPs). The results of this analysis
are summarised in Table~\ref{TAB_kernel_details}.

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    10      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
Working set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. $n$ is the problem size -- the length of one side of the
  square domain.}
\label{TAB_kernel_details}
\end{table}

We measure the mean time to execute this region of code as 0.0020
seconds (Intel v.16 on Haswell E5-1620 v2) for the 256 domain. This
gives us $\frac{108*256^2}{0.002} = 3.5$ GFLOPS. In contrast,
instrumenting this section of code with the LIKWID marker API and
using likwid-perfctr we measure 4.7 GFLOPS.

For the 128 the measured time per kernel call is 0.570033E-03 s.
$108*128^2/0.570e-3 = 3.1$ GFLOPS which is 25\% less than we measure
directly with likwid.  We can use the value reported by likwid to get
the FLOP count: 4.269 GFLOPS $\times 1.677 s = 7.1615217$ GFLOPs. This is for
3000 steps of $128^2$ so one kernel call is: $7.1615217/(3000*128^2) =
145.7$ FLOPs. This means that $AI = 145.7*7/(31*64) = 0.514$ FLOP/byte

\begin{figure}
\centering
\includegraphics[width=120mm]{roofline_archer}
\caption{Comparison of the performance achieved by nemolite2d and
  Shallow on Archer (Intel Ivybridge)}
\label{FIG_roofline_archer}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=120mm]{roofline_haswell}
\caption{Comparison of the performance achieved by nemolite2d and
  Shallow on Intel Haswell}
\label{FIG_roofline_haswell}
\end{figure}

>>>>

We want to use the Roofline Model to establish the efficiency (or
otherwise) of two shallow-water benchmark codes that we have used in
the GOcean project. The first of these codes, Shallow, was originally
written by Paul Swarztrauber over 30 years ago and has been optimised
by various people over that time. In contrast, nemolite2d is a new,
un-optimised code that has been extracted from NEMO.

Although we have investigated how the performance of the \psykal
version of nemolite2d compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound.

In this report we investigate the application of the Roofline Model to
these two, serial, shallow-water codes.

@@

We have performed this investigation on two different Intel CPUs; an
E5-1620 v.2 and an E5-2697 v.2. Both of these use the Intel Ivy Bridge
microarchitecture and have Ivy Bridge-EP cores. More details are given
in Table~\ref{TAB_stream_linpack}.

\begin{table}
\begin{tabular}{|l|c|c|c|c|}
\hline
  \multicolumn{5}{|c|}{E5-1620 v.~2, 3.7GHz} \\
  \hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{28.32} \\
\hline
 &  L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB   & 256 KB   & 10 MB    &  --         \\
\hline
STREAM2 (GB/s)   &  117  &  65.8  & 46.7 &  20.5 \\
  \hline
\multicolumn{5}{|c|}{E5-2697 v.~2, 2.7GHz}  \\
\hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{25.41} \\
\hline
 & L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB  & 256 KB   & 30 MB    &  --         \\
\hline
STREAM2 (GB/s) & 108 & 54.0 & 37.7  &  17.6 \\
\hline
\end{tabular}
\caption{The STREAM2 (DAXPY) and LINPACK benchmark results for a
  single thread running on each of the two CPUs considered here.}
\label{TAB_stream_linpack}
\end{table}

@@

Interpreting the results of the STREAM benchmark in order to get
bandwidth figures for all levels of the cache hierarchy is not
straightforward. We therefore used the STREAM2 benchmark which is
intended for this purpose. In Figure~\ref{FIG_stream2} we plot the
results obtained from the DAXPY component of the STREAM2 benchmark on
both of the CPUs examined here. The plateaux in bandwidth for L2, L3
and main memory are clear but that for L1 is considerably noisier. The
final values we have used in the roofline plots are given in
Table~\ref{TAB_stream_linpack}.

@@

@@

Initially, we hoped to use Hardware Performance Counters (HWPCs) in
order to get an accurate measure of the FLOPS and memory traffic
associated with specific kernels in each of the models.

The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
access to HWPCs on Intel and AMD CPUs.
However, there are issues with the accuracy of these counters,
especially for FLOP-related measurements on Ivy Bridge architectures. A
tool is included in the likwid distribution that tests the accuracy of
HWPC-derived data by running micro-benchmarks for which the FLOP count
and memory traffic are known:
\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
footprint spills out of L1 cache (which all of our nemolite2d
benchmark configurations do), the performance counter results
significantly overestimate the FLOPs performed.

@@

Although the Continuity kernel is very simple, the u-Momentum kernel
is not and in particular, includes a call to the {\tt sin()} function.
In order to estimate the cost of this function in FLOPs, the following
code was compiled (with the Intel compiler and -O1) and run with
likwid-perfctr:
\begin{verbatim}
  my_sum = 0.0d0
  do i = 1, 1000
    my_sum = my_sum + sin(real(i))
  end do
  write (*,*) 'My sum = ', my_sum
\end{verbatim}
This measurement revealed that the {\tt sin()} function requires
approximately 13 FLOPs.

A key component of the Roofline model is the Arithmetic Intensity of
the code being executed:
\begin{equation}
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\end{equation}
Since HWPC data proved to be unreliable, we calculated this quantity
manually by examining the source code and counting the number of
memory references and arithmetic operations that it contained. In
doing this counting we assume that any references to adjacent array
elements are fetched in a single cache-line and thus only count once.

To illustrate this, we consider the first loop-nest in the
time-stepping section of the Shallow code. This is a doubly-nested
loop over {\tt i} and {\tt j} with {\tt i} innermost. The body of the
loop is:
\begin{verbatim}
CU(I+1,J) = .5d0*(P(I+1,J)+P(I,J))*U(I+1,J)
CV(I,J+1) = .5d0*(P(I,J+1)+P(I,J))*V(I,J+1)
Z(I+1,J+1) =(FSDX*(V(I+1,J+1)-V(I,J+1))- &
             FSDY*(U(I+1,J+1)-U(I+1,J)))/&
            (P(I,J)+P(I+1,J)+            &
             P(I+1,J+1)+P(I,J+1))
H(I,J) = P(I,J)+0.25d0*                   &
         (U(I+1,J)*U(I+1,J)+U(I,J)*U(I,J) & 
         +V(I,J+1)*V(I,J+1)+V(I,J)*V(I,J))
\end{verbatim}
This code writes to four memory locations ({\tt CU}, {\tt CV}, {\tt Z}
and {\tt H}) and reads from ten distinct locations.  However, if we
assume caching of reads from locations adjacent in memory ({\it e.g.} {\tt
  P(i,j)} and {\tt P(i+1,j)}) then we have only six distinct reads.
If we assume that each of these ten memory accesses will touch a
separate cache line then this code fragment will result in the moving
of ten cache lines, each of which is 64 bytes (eight double-precision
words). However, as noted above, this code fragment is the body of a
loop nest. Therefore, upon the next loop trip {\tt i} will have been
incremented by one and the majority of the accessed values will still
be in cache. In fact, a new cache line will only have to be fetched
every seven trips of the (innermost) loop and thus the mean memory
traffic per loop trip is approximately $640/7$ bytes.

We can also count the number of floating-point arithmetic operations
(FLOPs) in this code fragment which gives us 12 multiplications and 12
additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
kernel of nemolite2d then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to {\tt
  sin()}. There are 31 distinct memory references. This gives an $AI$
of approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we
assume the call to {\tt sin()} costs 13 FLOPs). The results of this
analysis are summarised in Table~\ref{TAB_kernel_details}.

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    10      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
Working-set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. $n$ is the problem size -- the length of one side of the
  square domain.}
\label{TAB_kernel_details}
\end{table}

We measure the mean time to execute the u-Momentum kernel over the
whole domain as 0.0020 seconds (Intel v.16 on the E5-1620) for the
$256^2$ problem size. This gives us a performance of
$\frac{108*256^2}{0.002} = 3.5$ GFLOPS. In contrast, instrumenting
this section of code with the LIKWID marker API and using
likwid-perfctr we measure 4.7 GFLOPS.

For the $128^2$ domain, the measured time is 0.570033E-03 s. This
gives a performance of $108*128^2/0.570e-3 = 3.1$ GFLOPS which is 25\%
less than we measure directly with likwid.

Using these figures we can finally construct the Roofline model for
the two CPUs and the various kernels. In
Figure~\ref{FIG_roofline_archer} we plot the model for the E5-2697 CPU
and show the performance of the 1st loop-nest of Shallow and the
u-Momentum kernel of nemolite2d.

\begin{figure}
\centering
\includegraphics[width=120mm]{roofline_archer}
\caption{Comparison of the performance achieved by nemolite2d and
  Shallow on the E5-2697}
\label{FIG_roofline_archer}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=120mm]{roofline_haswell}
\caption{Comparison of the performance achieved by different nemolite2d and
kernels on the E5-1620.}
\label{FIG_roofline_haswell}
\end{figure}
